---
title             : "EDLD 654 Final Project: BrainAGE in Adolescence"
shorttitle        : "Adolescent BrainAGE"

author: 
  - name          : "Lucy Whitmore"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author


affiliation:
  - id            : "1"
    institution   : "University of Oregon"



  
keywords          : "keywords"
wordcount         : "X"

bibliography      : ["r-references.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

All materials and scripts are available at https://github.com/LucyWhitmore/EDLD-654

# Research Problem
As both neuroimaging and computational methods have improved in recent years, there has been interest in using machine learning methods to identify deviations in normative brain aging and development. One proposed methods is the Brain-Age Gap Estimation (BrainAGE), which quantifies the difference between one's chronological age and their age as predicted by machine learning models trained on neuroimaging data, often structural MRI measures. In older adult populations a positive BrainAGE, or a predicted age older than someone's chronological age, has been interpreted as reflecting premature brain aging, and has been associated with risk for Alzheimer's disease and cognitive decline. More recently, there has been interest in applying BrainAGE models to adolescent populations, where it has been hypothesized that BrainAGE could be related to risk for psychopathologies such as anxiety and depression. However, the majority of BrainAGE models are either trained exclusively on data from adult populations, or use dataframe from across the lifespan, but are not clear about how many adolescents are actually included in the model training.


Creating a BrainAGE model specifically for adolescent populations could improve the quality of predictions, better enabling researchers to accurately predict age. By improving BrainAGE predictions, we can also establish which outcomes and processes are related to BrainAGE in adolescence, both improving our understanding of development, and providing a possible indicator of risk.  


# Data Description
The data used in the creation of the following models were generated for use in this project, and are designed to match the format and distributions of data from the Adolescent Brain and Cognitive Development Study, a longitudinal multisite study of nearly 11,000 adolescents from the US, who are followed for 10 years. Participants were recruited when they were between 9-10 years old, and participate in an MRI scan every two years. Currently, data are available from the first two waves of data collection. 

The data consists of XXX columns and XXX rows. One column represents age, expressed in months, which will be used as the outcome variable. The predictors are XXX numeric columns, each representing a volume, area, or thickness measurement from a specific brain region. Column names ending with _Vol represent volume measurements, columns ending with _Area represent area measurements, and columns ending with _Thick represent thickness measurements. All predictors are continuous variables. As the data were simulated, there were no missing values. During data preparation, predictors with zero or near-zero variance were removed, and all predictors were standardized.

The data were split into training and testing samples using an 80/20 split. 


# Model Description
To determine which model type provides the best fit, three different modeling approaches were used. These models included an unregularized linear regression, a linear regression with a lasso penalty, and a gradient boosting tree. All three models were constructed using 10-fold cross-validation, and were evaluated using MAE and RMSE, as these metrics are commonly used in the BrainAGE literature. Hyperparameter tuning procedures for each model are described below. 

## Model 1 - Unregularized Linear Regression

## Model 2 - Linear Regression with LASSO Penalty

## Model 3 - Gradient Boosting Tree

# Model Fit
The best fiting model was XXX, followed by XXX, then XXX. 

## Model 1

## Model 2

## Model 3

# Discussion and Conclusion




We used `r cite_r("r-references.bib")` for all our analyses.




\newpage

# References

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup
